---
title: "Econ 220B PS 2"
author: "AC Watt"
email: aaron@acwatt.net
date: "2021-12-03"
output:
  pdf_document:
    toc: true
    toc_depth: 2
header-includes:
   - \usepackage{dcolumn}
   - \usepackage{amsmath}
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatstyle{plaintop}
   - \restylefloat{table}
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'ARE_213_PS2b.pdf')) })
---

```{r setup, message=F, echo=F}
rm(list=ls())
knitr::opts_chunk$set(echo = F)
```
<!--
R version 3.6.3 (2020-02-29)
Purpose of script: run code necessary for ps2b for ARE 213

Notes: Need to open ps2b.Rmd from the ps2b folder (without Rstudio being started first)
if Rstudio is already started, the working directory will not be set
to the ps2b/ folder

\usepackage{dcolumn}: dcolumn is needed in latex compilation for multicolumn tables
-->

```{r Settings}
# stargazer table type (html, latex, or text)
# Change to latex when outputting to PDF, html when outputting to html
table_type = "latex"
Cache = TRUE

```




```{r packages, message=F, warning=F}
# Packages
library(stargazer)
library(ggplot2)
library(readxl)
library(tidyverse)
library(lubridate)
```



\newpage
<!--=========================================================================-->
# Part a
<!--=========================================================================-->
**Construct sample statistics similar to Tabel IIa and IIb on page 1003.**
\vspace{1em}


```{r Load Data, eval=T, cache=Cache}
# Load data from PS2a with previous log variables
f = 'rust_data_clean.xlsx'
headers = read_excel(f, sheet='header_names', col_names='header')
data = read_excel(f, 1)
sheets = excel_sheets(f)[1:8]

replacement_dates = read_excel(f, sheet=sheets[1], n_max=11, col_names=F, col_types="numeric") %>%
    t() %>%
    data.frame(stringsAsFactors = F) %>%
    mutate(sheet = sheets[1], 
           group = 1)

i = 2
for (sheet in sheets[2:8]) {
    df_ = read_excel(f, sheet=sheet, n_max=11, col_names=F, col_types="numeric") %>%
        t() %>%
        data.frame(stringsAsFactors = F) %>%
        mutate(sheet = sheet, 
               group = i)
    replacement_dates = rbind(replacement_dates, df_)
    i = i + 1
}

replacement_dates %>%
    rename(!!set_names(paste0("X", 1:11), headers$header))

replacement_dates = lapply(setNames(sheets, sheets), function(x) read_excel(f, sheet=x, n_max=11, col_names=F, col_types="numeric")) %>%
    bind_cols(., .id="Sheet") %>%
    data.frame()
row.names(replacement_dates) <- headers$header
replacement_dates %>%
    # mutate(headers = headers$header) %>%
    t() %>%
    data.frame(stringsAsFactors = F) %>%
    mutate(purchase_date = ymd(str_c(purchase.year, purchase.month, "-01")),
           replace1_date = ymd(str_c(engine.replacement.1.year, engine.replacement.1.month, "-01")),
           replace2_date = ymd(str_c(engine.replacement.2.year, engine.replacement.2.month, "-01")),
           months_to_replace1 = interval(purchase_date, replace1_date) %/% months(1),
           months_to_replace2 = interval(replace1_date, replace2_date) %/% months(1)) %>%
    select(purchase_date, replace1_date, months_to_replace1, replace2_date, months_to_replace2, everything())



for (sheet in sheets) {
    
}

month_odo = lapply(setNames(sheets, sheets), function(x) read_excel(f, sheet=x, skip=11)) %>%
    bind_cols(., .id="Sheet")
  
# attaching all dataframes together
data_frame = bind_rows(data_frame, .id="Sheet")
```




\newpage
<!--=========================================================================-->
# Part b
<!--=========================================================================-->
**Estimate his model using a two-step procedure. Please briefly document your
estimation procedure in words. Create a table similar to the first five columns
of Table IX on page 1021. If you forward simulate the value function, see
Pakes Pollard (1989) for reference on how to draw errors and compute confidence
intervals.**

















\newpage
<!--=========================================================================-->
# Part c
<!--=========================================================================-->
**Estimate his model using a nested fixed point procedure. Create a table
similar to the first five columns of Table IX on page 1021.**














\newpage
# Appendix A: R Code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```





















